\chapter{Basic SQL}

SQL (Structured Query Language) is used to interact with relational databases. It helps in retrieving, inserting, updating, and deleting data.

\section{Basic SQL Commands}

\begin{definitiontcb}{SQL}{def:sql}
SQL is a language for managing relational databases.
\end{definitiontcb}

\section{Creating a Database and Table}

To create a database and a table:

\begin{minted}{sql}
CREATE DATABASE mydatabase;

USE mydatabase;

CREATE TABLE students (
    id INT PRIMARY KEY,
    name VARCHAR(100),
    age INT
);
\end{minted}

\begin{exampletcb}{Creating a Students Table}{ex:students-table}
Creates `mydatabase' and `students' table with `id', `name', and `age' columns.
\end{exampletcb}

\section{Inserting Data}

To insert data into a table:

\begin{minted}{sql}
INSERT INTO students (id, name, age) VALUES (1, 'Alice', 22);
INSERT INTO students (id, name, age) VALUES (2, 'Bob', 24);
\end{minted}

\begin{exampletcb}{Inserting Data}{ex:insert-data}
Adds two student records into the `students' table.
\end{exampletcb}

\section{Retrieving Data}

To retrieve data from a table:

\begin{minted}{sql}
SELECT * FROM students;
SELECT name, age FROM students WHERE age > 22;
\end{minted}

\begin{exampletcb}{Retrieving Data}{ex:select-data}
Retrieves all records from `students' and selects names and ages of students older than 22.
\end{exampletcb}

\section{Updating Data}

To update records in a table:

\begin{minted}{sql}
UPDATE students SET age = 23 WHERE id = 1;
\end{minted}

\begin{exampletcb}{Updating Data}{ex:update-data}
Updates Alice's age to 23.
\end{exampletcb}

\section{Deleting Data}

To delete records from a table:

\begin{minted}{sql}
DELETE FROM students WHERE id = 2;
\end{minted}

\begin{exampletcb}{Deleting Data}{ex:delete-data}
Removes Bob's record from the `students' table.
\end{exampletcb}

\section{Filtering and Sorting}

To filter and sort data:

\begin{minted}{sql}
SELECT * FROM students WHERE age > 20 ORDER BY name ASC;
\end{minted}

\begin{exampletcb}{Filtering and Sorting}{ex:filter-sort}
Retrieves students older than 20 and sorts them alphabetically by name.
\end{exampletcb}
